---
title: "Megakaryocyte ROI incorporation"
author: "Geoff Ivison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(emconnect)
library(SpatialMap)
library(ggplot2)
library(sf)
```

# Introduction

The data for this project includes a number of megakaryocytes, which have very large and irregularly shaped nuclei. Deepcell/mesmer unfortunately performs pretty poorly and inconsistently at segmenting this cell type.

We created [a story in the visualizer](https://app.enablemedicine.com/share/vis/nI__XY) that marks each of these megakaryocytes in the dataset with a lasso ROI. While the original plan was to use these to create an annotation set and then "merge" these marked sub-cells into whole cells, we determined that using these ROIs as surrogate whole-cell segmentation masks was a more robust approach.

The goal of this notebook is to establish a proof of concept of an approach to incorporate the centroids of these megakaryocyte ROIs as custom "cells" into a SpatialMap object, to facilitate using the tools in the SpatialMap package for analyses that consider these megakaryocytes.



```{r}
# First, pulling in the dataset.
sm <- readRDS("sm_annotation_final_with_fatDroplets.RDS")
```


Pick only the most recent clustering based annotation set as a starting point. The labels of the new megakaryocyte objects will be appended to this annotation set.

```{r define global variables}
con <- emconnect()

FINAL_ANNO_NAME <- "anno10001_All_cells_combined_Final"

MEGAKARYOCYTE_STORY_NAME <- "Megakaryocyte"

DEMO_REGION <- "SchuerchCMLBM_c001_v001_r002_reg001"
```



```{r}
# Extracting the metadata from the `sm` object.
metadata <- projectMetadata(sm)
```


Pulling the ROI coordinates for the megakaryocyte story.

```{r}
megakaryocyte_ROIs <- get_stories(acquisition_ids = metadata$acquisition_id) %>%
  filter(story_title == MEGAKARYOCYTE_STORY_NAME)

# Some accidental ROIs with fewer than 3 points got added to the story
# filtering them out here

ROIs_at_least_three_coords <- megakaryocyte_ROIs %>%
  group_by(roi_id) %>%
  summarize(count = n()) %>%
  filter(count >= 3)

megakaryocyte_ROIs <- left_join(ROIs_at_least_three_coords, megakaryocyte_ROIs)
```

Plot a pair of nearby ROIs just to make sure it's looking right.

```{r}
# Pick one region
demo_region_ROIs <- megakaryocyte_ROIs %>%
  filter(acquisition_id == DEMO_REGION)

# Get super rough centroids by averaging each ROI. Not robust enough for later
# but good enough for picking two nearby ones

demo_region_ROIs <- demo_region_ROIs %>%
  group_by(roi_id) %>%
  summarize(avg_x = mean(x),
            avg_y = mean(y))

demo_ROIs_dist <- dist(data.frame(x = demo_region_ROIs$avg_x,
                                  y = demo_region_ROIs$avg_y))

close_pair <- which(as.matrix(demo_ROIs_dist)==min(demo_ROIs_dist),arr.ind=TRUE)

demo_ROI_IDs <- c(demo_region_ROIs$roi_id[close_pair[1,1]], demo_region_ROIs$roi_id[close_pair[1,2]])

demo_ROIs <- megakaryocyte_ROIs %>%
  filter(roi_id %in% demo_ROI_IDs) %>%
  # Just to put it closer to the origin
  mutate(x = x - min(x),
         y = y - min(y),
         roi_id = as.character(roi_id))

ggplot(demo_ROIs, aes(x = x, y = y, color = roi_id)) +
  geom_point()
```

# Identifying megakaryocyte coordinates

For each ROI, calculate the center of that ROI. Then split out by region for the next operation.

Will test this out on the demo ROIs pulled in the previous section.

```{r}
megak_as_sf <- demo_ROIs %>%
  st_as_sf(coords = c("x", "y")) %>%
  group_by(roi_id) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

# plot(megak_as_sf)

megak_centroids <- st_centroid(megak_as_sf) %>%
  st_coordinates() %>%
  as_tibble() %>%
  mutate(roi_id = megak_as_sf$roi_id) %>%
  rename(x = X, y = Y)

ggplot(demo_ROIs, aes(x = x, y = y, color = roi_id)) +
  geom_point() +
  geom_point(data = megak_centroids, shape = "x", size = 5)
```

Looks great! Will generalize this to the entire object.

Will start by defining a helper function.

```{r}
summarize_to_centroids <- function(df) {
  megak_as_sf <- df %>%
    st_as_sf(coords = c("x", "y")) %>%
    # Makes some assumptions about the df object
    group_by(roi_id) %>%
    summarise(geometry = st_combine(geometry)) %>%
    st_cast("POLYGON")
  
  suppressWarnings(megak_centroids <- st_centroid(megak_as_sf) %>%
                     st_coordinates() %>%
                     as_tibble() %>%
                     mutate(roi_id = megak_as_sf$roi_id,
                            # Want integer pixel coordinates
                            X = round(X),
                            Y = round(Y)) %>%
                     rename(centroid_x = X, centroid_y = Y))
  
  return(megak_centroids)
}
```


And then run that for each acquisition ID

```{r}
megakaryocyte_ROI_list <- megakaryocyte_ROIs %>%
  group_split(acquisition_id)

reg_names <- purrr::map_chr(megakaryocyte_ROI_list,
                            \(df) unique(df$acquisition_id))

names(megakaryocyte_ROI_list) <- reg_names

megakaryocyte_centroids <- purrr::map(megakaryocyte_ROI_list,
                                      summarize_to_centroids)
```

# Adding megakaryocyte coordinates into SpatialMap object

For each region, add the megakaryocyte centroids. Will have to fill in dummy data for the "Data", "NormalizedData", "ScaledData", and "bgData" slots--these values will be filled with zeroes.

Since this is a region-by-region operation, will start by defining a helper function and then use `.smapply`

```{r}
# Will use for expanding project metadata values to fill in cellMetadata
pm_cols <- colnames(projectMetadata(sm))

add_megakaryocytes <- function(region,
                               megakaryocyte_centroids) {
  acquisition_id <- region@id
  
  megakaryocytes <- megakaryocyte_centroids[[acquisition_id]]
  
  # If there aren't any megakaryocyte centroids to add, just return the unmodified region
  if (is.null(megakaryocytes)) return(region)
  
  base_cell.ids <- cells(region)[[1]]
  base_spatial <- embeddings(region, "spatial") %>%
    # Need to un-invert the y-axis to put these into 
    # the same coordinate space as the ROIs
    as_tibble() %>%
    mutate(y = facil::inv(y))
  base_dat <- Data(region, "Data")
  base_normdat <- Data(region, "NormalizedData")
  base_scaledat <- Data(region, "ScaledData")
  base_bg <- Data(region, "bgData")
  base_chInfo <- channelInfo(region)
  base_cm <- cellMetadata(region)
  base_features <- features(region)
  
  # Will start megakaryocyte centroid numbering at the highest cell number + 1
  max_base_cell_id <- max(base_cm$cell_id)
  num_megakaryocytes <- nrow(megakaryocytes)
  # Creating cell.ids
  megakaryocyte_ids <- seq(from = max_base_cell_id + 1,
                           to = max_base_cell_id + num_megakaryocytes)
  megakaryocyte.ids <- paste0(acquisition_id, ".", megakaryocyte_ids)
  # Creating spatial data slot
  # Leaving as a data frame for now so we can re-invert the coordinates
  # So plots will match the orientation in the visualizer
  megakaryocyte_spatial <- tibble(x = megakaryocytes$centroid_x,
                                  y = megakaryocytes$centroid_y)
  # Creating dummy data and bgData
  megakaryocyte_dat <- matrix(data = 0,
                              nrow = nrow(base_dat),
                              ncol = num_megakaryocytes)
  rownames(megakaryocyte_dat) <- rownames(base_dat)
  colnames(megakaryocyte_dat) <- megakaryocyte.ids
  megakaryocyte_bg <- matrix(data = 0,
                             nrow = nrow(base_bg),
                             ncol = num_megakaryocytes)
  rownames(megakaryocyte_bg) <- rownames(base_bg)
  colnames(megakaryocyte_dat) <- megakaryocyte.ids
  # Creating dummy cellMetadata
  # Will full join and fill in columns from project metadata
  megakaryocyte_cm <- tibble(cell.id = megakaryocyte.ids,
                             cell_id = megakaryocyte_ids) %>%
    mutate(!!FINAL_ANNO_NAME := "Megakaryocyte")
  # No changes to chInfo or features
  
  updated_cell.ids <- c(base_cell.ids, megakaryocyte.ids)
  # Re-inverting y coordinate so plots match visualizer orientation
  updated_spatial <- bind_rows(base_spatial, megakaryocyte_spatial) %>%
    mutate(y = facil::inv(y)) %>%
    as.matrix() %>%
    magrittr::set_rownames(updated_cell.ids)
  # cbind because these matrices are oriented differently than the spatial
  updated_dat <- cbind(base_dat, megakaryocyte_dat)
  updated_normdat <- cbind(base_normdat, megakaryocyte_dat)
  updated_scaledat <- cbind(base_scaledat, megakaryocyte_dat)
  updated_bg <- cbind(base_dat, megakaryocyte_dat)
  # Full join and then fill in the project metadata columns that were merged
  # All other columns in the cell metadata will be left as `NA`
  suppressMessages(
    updated_cm <- full_join(base_cm, megakaryocyte_cm) %>%
      tidyr::fill(all_of(pm_cols)) %>%
      magrittr::set_rownames(updated_cell.ids)
  )
  
  # Reassembling modified region object
  region_with_megakaryocytes <- createRegionObject(
    id = acquisition_id,
    coordinates = updated_spatial,
    Data = updated_dat,
    NormalizedData = updated_normdat,
    ScaledData = updated_scaledat,
    bgData = updated_bg,
    cellMetadata = updated_cm,
    channelInfo = base_chInfo,
    features = base_features,
    auto.spatial.representation = T
  )
  
  return(region_with_megakaryocytes)
}
```

With those values defined, running `.smapply`.

```{r}
sm <- .smapply(sm,
               add_megakaryocytes,
               megakaryocyte_centroids = megakaryocyte_centroids,
               parallel = TRUE,
               cores = 4,
               analyze = "regions")
```

Some validity checks.

Are the cell labels in place?

```{r}
cellMetadata(sm)[[FINAL_ANNO_NAME]] %>% unique()
```

Are the megakaryocytes in the right location spatially, and spatial representations are oriented correctly?

```{r fig.height=10, fig.width=10}
demo_RDL <- projectMetadata(sm) %>%
  filter(acquisition_id == DEMO_REGION) %>%
  pull(region_display_label)

plotRepresentation(sm[["SchuerchCMLBM_c001_v001_r002_reg001"]], "spatial",
                   what = FINAL_ANNO_NAME) +
  gghighlight::gghighlight(variable == "Megakaryocyte",
                           use_group_by = FALSE) +
  ggtitle(demo_RDL)
```

Looks good!

Saving the output.

```{r}
saveRDS(sm, file = "sm_annotation_final_with_fatDroplets_megakaryocytes.RDS")
```

***

# Session info

<details><summary>Click to expand</summary>

```{r session info, results='markup'}
sessionInfo()
```
</details>


